### 2022-11-27 dissertation/chapters/05-probability.tex
🚧 (final-review): Chapter 5 WIP, just before conjugate priors

\subsection{Beta probability distribution}\label{sec:probability:probability_distributions:beta}

@@

where $\Gamma$ is the Gamma function as defined in Equation~\eqref{eq:probability:probability_distributions:beta:gamma_func} below.

@@

It should be noted that the Gamma function can also be written as shown in Equation~\eqref{eq:probability:probability_distributions:beta:gamma_func_alt} below.

@@

The control parameters $\alpha$ and $\beta$, determine the shape of the distribution. There exists a special case where $\alpha = \beta$. This is referred to as the \textit{symmetric} \index{Beta probability distribution}Beta probability distribution. In the case where $\alpha = \beta = 1$ the distribution is equivalent to the uniform distribution over all points in its support. The \index{Beta probability distribution}Beta probability distribution for various values of $\alpha$ and $\beta$, including the symmetric version is presented in Figure \ref{fig:probability:probability_distributions:beta}.

@@

The expected value of $x$ is given below in Equation~\eqref{eq:probability:probability_distributions:beta:expected_value}.

@@

Similarly, the expected value of the natural logarithm of $x$ can be calculated as is shown below in Equation~\eqref{eq:probability:probability_distributions:beta:expected_value_ln}.

@@

In Equation~\eqref{eq:probability:probability_distributions:beta:expected_value_ln} above, $\psi$ is the logarithmic derivative of the Gamma function, called the \textit{Digamma} function. The Digamma function is given below in Equation~\eqref{eq:probability:probability_distributions:beta:digamma}.

@@

Finally, the mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:beta:mode} below.

@@

\subsection{Dirichlet probability distribution}\label{sec:probability:probability_distributions:dirichlet}

The \index{Dirichlet probability distribution}Dirichlet probability distribution is a family of multivariate continuous probability distributions over some $x$ in $K$ dimensions~\cite{ref:wackerly:2014}. The \index{Dirichlet probability distribution}Dirichlet probability distribution is the multivariate generalization of the \index{Beta probability distribution}Beta probability distribution and is thus sometimes referred to by its alternative name, the multivariate Beta probability distribution. The \index{Dirichlet probability distribution}Dirichlet probability distribution is parameterised by some vector $\vec{\alpha} = (\alpha_{1}$,  $\dots, \alpha_{K})$ $\forall_{k=1}^{K} \alpha_{k} > 0$, $\alpha_{k} \in \mathbb{R}$. Throughout this chapter, the vector notation is dropped for simplicity and readability. $\alpha$ is referred to as the concentration parameter. The \index{Dirichlet probability distribution}Dirichlet probability distribution of order $K \geq 2$ with parameters $\alpha$, denoted $Dir(\alpha)$, has a \acs{PDF} as is given in Equation~\eqref{eq:probability:probability_distributions:dirichlet:pdf} below.

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:pdf}
      P(x \vert \alpha) =  f_{Dir}(x; K, \alpha) = \frac{1}{B(\alpha)}  \prod_{k=1}^{K} x_{k}^{\alpha_{k} - 1}
\end{equation}

In Equation~\eqref{eq:probability:probability_distributions:dirichlet:pdf} above, the normalising constant $B(\alpha)$, is defined in Equation~\eqref{eq:probability:probability_distributions:dirichlet:norm_cost}.

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:norm_cost}
      B(\alpha) = \frac{\prod_{k=1}^{K} \Gamma(\alpha_{k})}{\Gamma(\alpha_{0})}
\end{equation}

In Equation~\eqref{eq:probability:probability_distributions:dirichlet:pdf} above, $\alpha_{0}$ is defined in Equation~\eqref{eq:probability:probability_distributions:dirichlet:alpha_0} below.

@@

Importantly, the set $\{x_{k}\}_{k=1}^{K}$ belongs to the standard $K-1$ probability simplex $S$, meaning that $x_{K} = 1 - \sum_{k=1}^{K-1}x_{k}$ with support $\forall_{k=1}^{K} x \in [0,1]$. Under the simplex $S$, this means that the sum over all values of the vector $\vec{x}$ must be 1. The simplex can thus be rewritten as $\sum_{k=1}^{K}x_{k} = 1$.

Similar to the \index{Beta probability distribution}Beta probability distribution, $\alpha$ determines the shape of the distribution in $K$ dimensions and thus, there also exist a special case, referred to as the \textit{symmetric} distribution when $\forall_{k=1}^{K} \alpha_{k} = c$, where $c$ is some constant. In the case where $c = 1$, the distribution is referred to as a \textit{flat} distribution and yields the uniform distribution over all points in $S$. The \index{Dirichlet probability distribution}Dirichlet probability distribution of order $K = 3$, for various values of $\alpha$, including the symmetric version is presented in Figure \ref{fig:probability:probability_distributions:dirichlet}.

\begin{figure}[htb]
      \centering
      \includegraphics[width=0.9\textwidth]{images/dirichlet.pdf}
      \caption{The \acfp{PDF} for the \index{Dirichlet probability distribution}Dirichlet probability distribution over the 2-simplex. The concentration parameter $\alpha$ is varied. The values of the \acs{PDF} are shown by the color maps with contour lines at equal values as indicated in the color bars~\cite{ref:dirichlet:2020}.}
      \label{fig:probability:probability_distributions:dirichlet}
\end{figure}

The expected value of $x$ can be calculated as shown in Equation~\eqref{eq:probability:probability_distributions:dirichlet:expected_value} below.

@@

Similarly, the expected value of the natural logarithm of $x_{k}$ can be calculated as defined in Equation~\eqref{eq:probability:probability_distributions:dirichlet:expected_value_ln} below.

@@

In Equation~\eqref{eq:probability:probability_distributions:dirichlet:expected_value_ln} above, $\psi$ is the Digamma function as defined in Equation~\eqref{eq:probability:probability_distributions:beta:digamma}. Finally, the mode of the distribution is given in Equation~\eqref{eq:probability:probability_distributions:dirichlet:mode} below.

@@

\subsection{Bernoulli probability distribution}\label{sec:probability:probability_distributions:bernoulli}

The \index{Bernoulli probability distribution}Bernoulli probability distribution is a discrete probability distribution over some random variable $x$ that takes the value of $1$ with probability $\theta$ and $0$ with probability $1-\theta$ ~\cite{ref:wackerly:2014}. The \index{Bernoulli probability distribution}Bernoulli probability distribution is denoted as $Ber(\theta)$ with support $x \in \{0, 1\}$. In probability theory, the \index{Bernoulli probability distribution}Bernoulli probability distribution is often used to explain the possible outcomes of a single experiment that asks a \textit{yes-no} question such as flipping a coin. The outcome of such an experiment is a Boolean value. The \index{Bernoulli probability distribution}Bernoulli probability distribution has a \acf{PMF}, given as

@@

\subsection{Binomial probability distribution}\label{sec:probability:probability_distributions:bin}

The \index{Binomial probability distribution}Binomial probability distribution is a discrete probability distribution over a random variable $x$ taking on a number of successes in $N$ sequential independent experiments that each ask a \textit{yes-no} question~\cite{ref:wackerly:2014}. The probability of a single independent experiment yielding a success is given as $\theta$ and the \index{Binomial probability distribution}Binomial probability distribution is denoted as $Bin(N, \theta)$, with support $x \in \{0, 1, \dots, N\}$.  It should be noted that the \index{Binomial probability distribution}Binomial probability distribution is an extension of the \index{Bernoulli probability distribution}Bernoulli probability distribution over $N$ independent sequential experiments, and thus each experiment also yields some Boolean outcome. When $N=1$, the experiment is referred to as a Bernoulli trial, and the distribution is just a \index{Bernoulli probability distribution}Bernoulli probability distribution. When $N > 1$, the sequence of outcomes is referred to as a Bernoulli process. The \acf{PMF} of the \index{Binomial probability distribution}Binomial probability distribution is given as

@@

The \index{Categorical probability distribution}Categorical probability distribution is a discrete probability distribution over some random variable, $x$, taking on any one of $K$ possible categories~\cite{ref:wackerly:2014}. There is no ordering to these categories and therefore, for simplicity, each category is assigned a numerical representative value such that $k \in \{1, 2, \dots, K\}$. The probabilities for all outcomes is given by the probability vector $\vec{\theta} = (\theta_{1}, \dots, \theta_{K})$.  This means that the probability $P(x=k)=\theta_{k}$ has support $x \in \{1, \dots, K\}$. The \index{Categorical probability distribution}Categorical probability distribution, denoted $Cat(\theta)$, is a generalization of the \index{Bernoulli probability distribution}Bernoulli probability distribution and is sometimes referred to it by its alternative names, i.e. the generalised \index{Bernoulli probability distribution}Bernoulli probability distribution or the Multinoulli probability distribution. In probability theory, the \index{Categorical probability distribution}Categorical probability distribution is often used to explain the outcome of a single experiment with more than two possible outcomes, such as rolling a six-sided die~\cite{ref:wackerly:2014}. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:pmf}
      P(x \vert \theta; K) = f_{Cat}(x; K, \theta) = \prod_{k=1}^{K}\theta_{k}^{[x = k]}
\end{equation}

where the term $[x = k]$ is the \textit{Iversion Bracket}~\cite{ref:iverson:1962}, yielding 1 if $x = k$ and 0 otherwise. Given class $k$, the categorical distribution simply yields $\theta_{k}$ as is follows:

@@

The \index{Multinomial probability distribution}Multinomial probability distribution is a discrete probability distribution over some random variable $x = (x_{1}, \dots\, x_{K})$ that takes on the counts for each occurrence of $K$ possible classes in $N$ independent trials~\cite{ref:wackerly:2014}. The probabilities for all possible outcomes in a single trial is given by the probability vector $\vec{\theta} = (\theta_{1}, \dots, \theta_{K})$. The \index{Multinomial probability distribution}Multinomial probability distribution, denoted $Mul(N, K, \theta)$, is thus the generalization of the \index{Binomial probability distribution}Binomial probability distribution to $K$ dimensions. Consider the following special cases.

@@

The support for the Multinomial is $\forall_{i=1}^{K} x_{k} \in \{1, \dots, N\}, \sum_{k=1}^{K}x_{k} = N$ and the \acs{PMF} for the \index{Multinomial probability distribution}Multinomial probability distribution is given below in Equation~\eqref{eq:probability:probability_distributions:multinomial:pmf}.

\begin{equation}
      \label{eq:probability:probability_distributions:multinomial:pmf}
      P(x \vert \theta; N; K) = f_{Mul}(x; N, K, \theta) = \frac{N!}{\prod_{k=1}^{K}x_{k}!} \prod_{k=1}^{K}\theta_{k}^{x_{k}}
\end{equation}

Similar to the \index{Categorical probability distribution}Categorical probability distribution, the random variable $x$ can also be encoded in binary format, yielding an $N \times K$ matrix $X$ of \index{Bernoulli probability distribution}Bernoulli probability distributions. The support is then given as $X \in \{0, 1\}^{N \times K}, \forall_{i=1}^{N}\sum_{k=1}^{K} x_{i,k} = 1$ so that the standard $K-1$ probability simplex $S$ still holds for each trial. The \acs{PMF} of the \index{Multinomial probability distribution}Multinomial probability distribution can then be rewritten as is presented in Equation~\eqref{eq:probability:probability_distributions:multinomial:pmf_alt} below.

\begin{equation}
      \label{eq:probability:probability_distributions:multinomial:pmf_alt}
      \begin{split}
            f_{Mul}(x; N, K, \theta) &= \frac{N!}{\prod_{k=1}(\sum_{i=1}^{N}x_{i, k})!}\prod_{i=1}^{N}\prod_{k=1}^{K}\theta_{k}^{\mathbbm{1}_{1}(x_{i,k})} \\
            &= \frac{N!}{\prod_{k=1}(\sum_{i=1}^{N}x_{i, k})!} \prod_{k=1}^{K}\theta_{k}^{\sum_{i=1}^{N}\mathbbm{1}_{1}(x_{i,k})} \\
            &= \frac{N!}{\prod_{k=1}(\sum_{i=1}^{N}x_{i, k})!} \prod_{k=1}^{K}\theta_{k}^{N_{k}} \\
      \end{split}
\end{equation}

In Equation~\eqref{eq:probability:probability_distributions:multinomial:pmf_alt} above, $N_{k}$ is a summary variable, denoting the number of times a category $k$ occurs over all trials in $N$.

The \index{Categorical probability distribution}Categorical and \index{Multinomial probability distribution}Multinomial probability distributions are presented as binary encoded vectors in order to simplify the proof of their conjugate priors as is shown in the following section.

>>>>

\subsection{Beta Probability Distribution}\label{sec:probability:probability_distributions:beta}

@@

where $\Gamma$ is the Gamma function defined as

@@

It should be noted that the Gamma function can also be written as

@@

The control parameters $\alpha$ and $\beta$ determine the shape of the distribution. There exists a special case where $\alpha = \beta$. This is referred to as the \textit{symmetric} \index{Beta probability distribution}Beta probability distribution. In the case where $\alpha = \beta = 1$ the distribution is equivalent to the uniform distribution over all points in its support. The \index{Beta probability distribution}Beta probability distribution for various values of $\alpha$ and $\beta$, including the symmetric version is illustrated in Figure \ref{fig:probability:probability_distributions:beta}.

@@

The expected value of $x$ is given as

@@

Similarly, the expected value of the natural logarithm of $x$ is calculated as

@@

where $\psi$ is the logarithmic derivative of the Gamma function, called the \textit{Digamma} function. The Digamma function is given as

@@

Finally, the mode of the distribution is given as

@@

\subsection{Dirichlet Probability Distribution}\label{sec:probability:probability_distributions:dirichlet}

The \index{Dirichlet probability distribution}Dirichlet probability distribution is a family of multivariate continuous probability distributions over some $\boldsymbol{x}$ in $K$ dimensions~\cite{ref:wackerly:2014}. The \index{Dirichlet probability distribution}Dirichlet probability distribution is the multivariate generalization of the \index{Beta probability distribution}Beta probability distribution and is thus sometimes referred to by its alternative name, i.e. the multivariate Beta probability distribution. The \index{Dirichlet probability distribution}Dirichlet probability distribution is parameterised by some vector $\boldsymbol{\alpha} = (\alpha_{1}$,  $\dots, \alpha_{K}), \forall_{k=1}^{K} \alpha_{k} > 0, \alpha_{k} \in \mathbb{R}$. The parameter, $\boldsymbol{\alpha}$, is referred to as the concentration parameter. The \index{Dirichlet probability distribution}Dirichlet probability distribution of order $K \geq 2$ with parameter $\boldsymbol{\alpha}$, denoted $Dir(\boldsymbol{\alpha})$, has a \acs{PDF} given as

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:pdf}
      P(\boldsymbol{x} \vert \boldsymbol{\alpha}) =  f_{Dir}(\boldsymbol{x}; K, \boldsymbol{\alpha}) = \frac{1}{B(\boldsymbol{\alpha})}  \prod_{k=1}^{K} x_{k}^{\alpha_{k} - 1}
\end{equation}

The normalising constant, $B(\boldsymbol{\alpha})$, is defined as

\begin{equation}
      \label{eq:probability:probability_distributions:dirichlet:norm_cost}
      B(\boldsymbol{\alpha}) = \frac{\prod_{k=1}^{K} \Gamma(\alpha_{k})}{\Gamma(\alpha_{0})}
\end{equation}

where $\alpha_{0}$ is defined as

@@

Importantly, the set $\{x_{k}\}_{k=1}^{K}$ belongs to the standard $K-1$ probability simplex $S$, meaning that $x_{K} = 1 - \sum_{k=1}^{K-1}x_{k}$ with support $\forall_{k=1}^{K} x_{k} \in [0,1]$. Under the simplex $S$, this means that the sum over all values of the vector $\boldsymbol{x}$ must be 1. The simplex can thus be rewritten as $\sum_{k=1}^{K}x_{k} = 1$.

Similar to the \index{Beta probability distribution}Beta probability distribution, $\boldsymbol{\alpha}$ determines the shape of the distribution in $K$ dimensions and thus, there also exist a special case, referred to as the \textit{symmetric} distribution when $\forall_{k=1}^{K} \alpha_{k} = c$, where $c$ is some constant. In the case where $c = 1$, the distribution is referred to as a \textit{flat} distribution and yields the uniform distribution over all points in $S$. The \index{Dirichlet probability distribution}Dirichlet probability distribution of order $K = 3$ for various values of $\boldsymbol{\alpha}$, including the symmetric version, is presented in Figure \ref{fig:probability:probability_distributions:dirichlet}.

\begin{figure}[htb]
      \centering
      \includegraphics[width=0.9\textwidth]{images/dirichlet.pdf}
      \caption{The \acfp{PDF} for the \index{Dirichlet probability distribution}Dirichlet probability distribution over the 2-simplex. The concentration parameter $\boldsymbol{\alpha}$ is varied. The values of the \acs{PDF} are shown by the color maps with contour lines at equal values as indicated in the color bars~\cite{ref:dirichlet:2020}.}
      \label{fig:probability:probability_distributions:dirichlet}
\end{figure}

The expected value of $x_{k}$ is calculated as

@@

Similarly, the expected value of the natural logarithm of $x_{k}$ is calculated as

@@

where $\psi$ is the Digamma function as defined in Equation~\eqref{eq:probability:probability_distributions:beta:digamma}. Finally, the mode of the distribution is given as

@@

\subsection{Bernoulli Probability Distribution}\label{sec:probability:probability_distributions:bernoulli}

The \index{Bernoulli probability distribution}Bernoulli probability distribution is a discrete probability distribution over some random variable $x$ that takes the value of $1$ with probability $\theta$ and $0$ with probability $1-\theta$ ~\cite{ref:wackerly:2014}. The \index{Bernoulli probability distribution}Bernoulli probability distribution is denoted as $Ber(\theta)$ with support $x \in \{0, 1\}$. In probability theory, the \index{Bernoulli probability distribution}Bernoulli probability distribution is often used to explain the possible outcomes of a single experiment that asks a \textit{yes-no} question such as flipping a coin. The outcome of such an experiment is a Boolean value. The \index{Bernoulli probability distribution}Bernoulli probability distribution has a \acs{PMF}, given as

@@

\subsection{Binomial Probability Distribution}\label{sec:probability:probability_distributions:bin}

The \index{Binomial probability distribution}Binomial probability distribution is a discrete probability distribution over a random variable $x$ taking on a number of successes in $N$ sequential independent experiments that each ask a \textit{yes-no} question~\cite{ref:wackerly:2014}. The probability of a single independent experiment yielding a success is given as $\theta$ and the \index{Binomial probability distribution}Binomial probability distribution is denoted as $Bin(N, \theta)$, with support $x \in \{0, 1, \dots, N\}$.  It should be noted that the \index{Binomial probability distribution}Binomial probability distribution is an extension of the \index{Bernoulli probability distribution}Bernoulli probability distribution over $N$ independent sequential experiments, and thus each experiment also yields some Boolean outcome. When $N=1$, the experiment is referred to as a Bernoulli trial, and the distribution is just a \index{Bernoulli probability distribution}Bernoulli probability distribution. When $N > 1$, the sequence of outcomes is referred to as a Bernoulli process. The \acs{PMF} of the \index{Binomial probability distribution}Binomial probability distribution is given as

@@

The \index{Categorical probability distribution}Categorical probability distribution is a discrete probability distribution over some random variable, $x$, taking on any one of $K$ possible categories~\cite{ref:wackerly:2014}. There is no ordering to these categories and therefore, for simplicity, each category is assigned a numerical representative value such that $k \in \{1, 2, \dots, K\}$. The probabilities for all outcomes is given by the probability vector $\boldsymbol{\theta} = (\theta_{1}, \dots, \theta_{K})$.  This means that the probability $P(x=k)=\theta_{k}$ has support $x \in \{1, \dots, K\}$. The \index{Categorical probability distribution}Categorical probability distribution, denoted $Cat(\boldsymbol{\theta})$, is a generalization of the \index{Bernoulli probability distribution}Bernoulli probability distribution and is sometimes referred to it by its alternative names, i.e. the generalised \index{Bernoulli probability distribution}Bernoulli probability distribution or the Multinoulli probability distribution. In probability theory, the \index{Categorical probability distribution}Categorical probability distribution is often used to explain the outcome of a single experiment with more than two possible outcomes, such as rolling a six-sided die~\cite{ref:wackerly:2014}. The \acs{PMF} of the \index{Categorical probability distribution}Categorical probability distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:categorical:pmf}
      P(x \vert \boldsymbol{\theta}; K) = f_{Cat}(x; K, \boldsymbol{\theta}) = \prod_{k=1}^{K}\theta_{k}^{[x = k]}
\end{equation}

where the term, $[x = k]$, is the \textit{Iversion Bracket}~\cite{ref:iverson:1962}, yielding 1 if $x = k$ and 0 otherwise. Given class $k$, the categorical distribution simply yields $\theta_{k}$ as follows:

@@

The \index{Multinomial probability distribution}Multinomial probability distribution is a discrete probability distribution over some random variable $\boldsymbol{x} = (x_{1}, \dots\, x_{K})$ that takes on the counts for each occurrence of $K$ possible classes in $N$ independent trials~\cite{ref:wackerly:2014}. The probabilities for all possible outcomes in a single trial is given by the probability vector, $\boldsymbol{\theta} = (\theta_{1}, \dots, \theta_{K})$. The \index{Multinomial probability distribution}Multinomial probability distribution, denoted $Mul(N, K, \boldsymbol{\theta})$, is thus a generalization of the \index{Binomial probability distribution}Binomial probability distribution to $K$ dimensions. Consider the following special cases:

@@

The support for the Multinomial is $\forall_{i=1}^{K} x_{k} \in \{1, \dots, N\}, \sum_{k=1}^{K}x_{k} = N$ and the \acs{PMF} for the \index{Multinomial probability distribution}Multinomial probability distribution is given as

\begin{equation}
      \label{eq:probability:probability_distributions:multinomial:pmf}
      P(\boldsymbol{x} \vert \boldsymbol{\theta}; N; K) = f_{Mul}(x; N, K, \boldsymbol{\theta}) = \frac{N!}{\prod_{k=1}^{K}x_{k}!} \prod_{k=1}^{K}\theta_{k}^{x_{k}}
\end{equation}

Similar to the \index{Categorical probability distribution}Categorical probability distribution, the random variable $x$ can also be encoded in binary format, yielding an $N \times K$ matrix $X$ of \index{Bernoulli probability distribution}Bernoulli probability distributions. The support is then given as $\boldsymbol{X} \in \{0, 1\}^{N \times K}, \forall_{i=1}^{N}\sum_{k=1}^{K} x_{i,k} = 1$ so that the standard $K-1$ probability simplex $S$ still holds for each trial. The \acs{PMF} of the \index{Multinomial probability distribution}Multinomial probability distribution is then rewritten as

\begin{equation}
      \label{eq:probability:probability_distributions:multinomial:pmf_alt}
      \begin{split}
            f_{Mul}(\boldsymbol{X}; N, K, \boldsymbol{\theta}) &= \frac{N!}{\prod_{k=1}(\sum_{i=1}^{N}x_{i, k})!}\prod_{i=1}^{N}\prod_{k=1}^{K}\theta_{k}^{\mathbbm{1}_{1}(x_{i,k})} \\
            &= \frac{N!}{\prod_{k=1}(\sum_{i=1}^{N}x_{i, k})!} \prod_{k=1}^{K}\theta_{k}^{\sum_{i=1}^{N}\mathbbm{1}_{1}(x_{i,k})} \\
            &= \frac{N!}{\prod_{k=1}(\sum_{i=1}^{N}x_{i, k})!} \prod_{k=1}^{K}\theta_{k}^{N_{k}} \\
      \end{split}
\end{equation}

where $N_{k}$ is a summary variable, denoting the number of times a category $k$ occurs over all trials in $N$.
